\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-12-18 doc/publications/plos16/plos16.tex
added reference and changes to plos16.tex

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the {\sc randomForest}
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a very poor fit and still have an accurate ranking of variable
  importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth}
    to 4 in their classification example. This is wrong. The
    individual trees should be grown to a large size to ensure that
    they give an unbiased estimate.}.
  In the TCGA example the RandomForests code produced trees with depths of 30+. 
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
In this section we demonstrate two features of \cursedforest\, variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000
samples, as discussed in the method section.

After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
classification error and the 20 most important variables.

Fig.~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
(\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
15 remaining variables from the list.

Fig.~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB
(out of bag) classification error.
By $mtry=0.8$ the error is converging to X, which is due to Y.
\marginpar{what are ``X'' and ``Y''. Piotr}

@@

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error-prod.png}
    \caption{OOB classification error} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/rbo-prod.png}
    \caption{The  rank-biased-overlap measure.} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
    as a function of \mtry and \ntree. {\bf a0} OOB (out of bag) classification error.  {\bf b)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
  \label{figure:synth}
\end{adjustwidth}
\end{figure}

@@

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
                                                & RF - Spark ML  &  0.05           &           7 min 44 seconds         \\
                                                & RF - R         & 0.05       & 24 min           \\
                                                & ranger - R      & 0.06       & 1 min 10 seconds       \\
                                                & ranger - c++     & 0.06       & 7 min            \\

 \hline
\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  \\
                                                    & RF - R        & NA         & NA                \\
                                                    & H2O           & 0.05       & 16 hours         \\
                                                    & ranger - R    & 0.06       & 4 min 30 seconds     \\
\hline
\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min         \\
                                                    & RF - Spark ML  & 0.04       & 8hr 29min        \\
                                                    & ranger - R      &  0.05       & 28 mins         \\
\hline
\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min               \\
                                                    & RF - Spark ML  &    NA         &   NA               \\
                                                    & ranger - R      & 0.03      &         \\
& ranger - R        &             0.973     &      03:58           \\
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX  & 7hr 14min \\
& Spark ML & NA & NA  \\
& ranger (C++)       &        NA     &        NA      \\
& ranger (C++)       &        NA     &        NA    \\
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{timingtable}
%\end{adjustwidth}
\end{minipage}
\end{table}

>>>>

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the {\sc randomForest}
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a very poor fit and still have an accurate ranking of variable
  importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  {\sc randomForest} by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
In this section we demonstrate two features of \cursedforest\, a variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000
samples, as discussed in the methods section.

After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
classification error and the rank-biased-overlap
\cite[]{Webber.et.al.2010} measure.

% Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} illustrates that increasing the number of predictors sampled for splitting at each node
% (\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
% the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
% measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
% difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
% with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
% 15 remaining variables from the list.

% Fig.~\ref{figure:rbo-prod.png} illustrates that increasing \mtry\ also improves the OOB
% (out of bag) classification error.
% By $mtry=0.8$ the error is converging to X, which is due to Y.
% \marginpar{what are ``X'' and ``Y''. Piotr}

% From the synthetic data we can conclude that \cursedforest\ is able to perform accurate feature selection as well as
% accurate classification for large numbers of samples with large feature vectors.

Fig.~\ref{figure:synth1} illustrates the effect of varying  \mtry\ and \ntree\ on these two measures. 
Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows that depends strongly on \ntree
whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the rbo measure is more dependent on \mtry.

@@

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error-prod.png}
    \caption{OOB classification error} 
    \label{figure:out-of-bag-prediction-error-prod.png} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/rbo-prod.png}
    \caption{The  rank-biased-overlap measure.} 
    \label{figure:rbo-prod.png} 
    \vspace{4ex}
  \end{subfigure} 
  \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
    as a function of \mtry\ and \ntree. Note that the plot colors have been chose so that the lighter color indicates
    the better result, that is, a lower OOB and a higher RBO.}
  \label{figure:synth1}
\end{adjustwidth}
\end{figure}

@@

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
                                                & RF - Spark ML  &  0.05           &           7 min 44 seconds         \\
                                                & randomForest - R         & 0.05       & 24 min           \\
                                                & ranger - R      & 0.06       & 1 min 10 seconds       \\
                                                & ranger - c++     & 0.06       & 7 min            \\

 \hline
\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548}   &               &            &                  \\
                                                    & randomForest - R      & NA         & NA                \\
                                                    & H2O           & 0.05       & 16 hours         \\
                                                    & ranger - R    & 0.06       & 4 min 30 seconds     \\
\hline
\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min         \\
                                                    & RF - Spark ML  & 0.04       & 8hr 29min        \\
                                                    & ranger - R      &  0.05       & 28 mins         \\
\hline
\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min               \\
                                                    & RF - Spark ML  &    NA         &   NA               \\
                                                    & ranger - R      & 0.03      &         \\
                                                    & ranger - R        &             0.973     &      03:58           \\
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX  & 7hr 14min \\
& Spark ML & NA & NA  \\
& ranger (C++)       &        NA     &        NA      \\
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{timingtable}
%\end{adjustwidth}
\end{minipage}
\end{table}
